### 2020-05-11 README.Rmd
Updated README.Rmd with author information
Second version of paper after first round of anonymous reviews.

@@

In 2020 Google released a set of Community Mobility Reports (GCMR). These reports are based on the company's location-tracking capabilities and measure changes in mobility with respect to a baseline. This novel source of data offers an opportunity to investigate the way mobility potentially correlates with transmission of COVID-19. Using data from the New York Times on COVID-19 cases and GCMR, this paper presents an analysis of mobility levels and new daily reported cases of COVID-19 by state in the US. The results provide insights about the utility and interpretability of GCMR for COVID-19 research and decision-making.

@@

The main policy tool to control the spread of the COVID-19 pandemic has been stay-at-home orders. Concurrently, numerous efforts exist to track the progress and the impact of the pandemic, and new sources of data include the recently-released Google Community Mobility Reports [GCMR](https://www.google.com/covid19/mobility/), as well as The New York Times repository of [COVID-19 data](https://github.com/nytimes/covid-19-data). These two open data sets offer novel opportunities to investigate in quasi-real time the relationship between mobility patterns and transmission of COVID-19.

@@

The source document for this paper is an `R` markdown document available from the following repository:

https://github.com/paezha/Google-Mobility-Reports-and-COVID-19-US/tree/fcd055878020bf567cd4deabb16c8056189eb45c/Covid-19-Google-CMR-US

@@

GCMR use aggregated and anonymized data to chart changes in mobility with respect to different classes of places, namely retail and recreation, groceries and pharmacies, parks, transit stations, workplaces, and residential. Mobility indicators for each of these places are reported as percentage changes from a baseline level, which corresponds to the median value of mobility of identical days of the week during the period between January 3 and Feb 6, 2020. Covid-19 data is compiled by The New York Times based on reports from state and local health agencies.

For analysis, all mobility indicators are centered so that the value of 1 is the base mobility, and a 0.01 deviation corresponds to a 1% change. The incubation time of the disease has been estimated to be between 2 and 12 days (95% interval) by Lauer et al. [-@Lauer2020incubation]. Given this, it is to be expected that any changes in mobility will have a lagged effect on the discovery of new cases. For this reason, lagged moving averages of the mobility indicators are calculated. Furthermore, it is possible that mobility and reports of new cases of COVID-19 are endogenous, if the public adjust their mobility according to reports of the incidence. Therefore, in addition to being consistent with an incubation period, use of lagged indicators also helps to break this potential endogeneity.

```{r descriptive-statistics, echo=FALSE}
data.frame(Variable = c("New Cases",
                        "date",
                        "retail",
                        "groceries",
                        "parks",
                        "transit",
                        "work",
                        "residential"),
           Definition = c("New Daily Cases of COVID-19",
                        "Date",
                        "Retail and Recreation",
                        "Groceries and Pharmacies",
                        "Parks",
                        "Transit Stations",
                        "Workplaces",
                        "Residential"),
           min = c(as.character(min(c19$new_cases)),
                   as.character(min(c19$date)),
                   as.character(round(min(c19$retail), 2)),
                   as.character(round(min(c19$groceries), 2)),
                   as.character(round(min(c19$parks), 2)),
                   as.character(round(min(c19$transit), 2)),
                   as.character(round(min(c19$work), 2)),
                   as.character(round(min(c19$residential), 2))),
           median = c(as.character(median(c19$new_cases)),
                   as.character(median(c19$date)),
                   as.character(round(median(c19$retail), 2)),
                   as.character(round(median(c19$groceries), 2)),
                   as.character(round(median(c19$parks), 2)),
                   as.character(round(median(c19$transit), 2)),
                   as.character(round(median(c19$work), 2)),
                   as.character(round(median(c19$residential), 2))),
           max = c(as.character(max(c19$new_cases)),
                   as.character(max(c19$date)),
                   as.character(round(max(c19$retail), 2)),
                   as.character(round(max(c19$groceries), 2)),
                   as.character(round(max(c19$parks), 2)),
                   as.character(round(max(c19$transit), 2)),
                   as.character(round(max(c19$work), 2)),
                   as.character(round(max(c19$residential), 2))),
           sd = c(as.character(round(sd(c19$new_cases), 2)),
                   as.character(""),
                   as.character(round(sd(c19$retail), 2)),
                   as.character(round(sd(c19$groceries), 2)),
                   as.character(round(sd(c19$parks), 2)),
                   as.character(round(sd(c19$transit), 2)),
                   as.character(round(sd(c19$work), 2)),
                   as.character(round(sd(c19$residential), 2)))) %>%
  kable(#"latex",
        "html",
        booktabs = TRUE,
        caption = "\\label{tab:descriptive-statistics}Descriptive statistics of the data set") %>%
  kable_styling(latex_options = c("striped", "scale_down", "hold_position")) %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

The lagged indicators are calculated as the mean of the mobility indicator using the values from date-minus-12-days to date-minus-2-days. Furthermore, using the cumulative number of reported COVID-19 cases, the total number of new daily cases is calculated. This variable (log-transformed after adding a small constant) is paired to the corresponding lagged moving average of the mobility indicators. The log-transformation is useful to avoid negative values of daily new cases when making predictions. Analysis is based on correlation analysis, multivariate regression, and data visualization. Table \ref{tab:descriptive-statistics} shows the descriptive statistics of the data set. Temporal coverage is from February 27 to April 26, 2020. The maximum number of new daily cases during this period is 12,126.

@@

Table \ref{tab:correlation-analysis} shows that the mobility indicators are highly correlated with each other. This is not surprising: it is well-known that mobility involves time-use trade-offs: the more there is of one kind of mobility, the less time there is available for any other. That said, the strongest correlation with the outcome of interest is for residential-based mobility. To avoid multicollinearity, parks-related mobility is selected as a covariate since it has the lowest correlation with residential-based mobility.

```{r check-correlations, echo=FALSE}
#Check the correlation between the mobility indicators and incidence:
c19 %>% 
  select(log_new_cases, retail, groceries, parks, transit, work, residential) %>%
  cor(use = "pairwise.complete.obs") %>%
  kable(#"latex",
        "html",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:correlation-analysis}Simple correlation between log(New Cases) and the mobility indicators")  %>%
  #kable_styling(latex_options = c("striped", "scale_down", "hold_position")) %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

A regression model uses the log of new daily cases as the dependent variable. The covariates are parks- and residential-related mobility, which enter the regression in the form of a second order polynomial expansion. In addition, the date is introduced to account for the temporal trend of the pandemic. Finally, an indicator variable for the state of New York and an interaction with the date are used to distinguish the unusually high incidence of the disease there. The results appear in Table \ref{tab:model-results}. The model provides a good fit to the data and all variables reported are significant at $p<0.05$ or better.

There is an overall temporal trend that indicates a growing number of cases over time, but at a decreasing rate (see negative sign of date^2). New York has on average more new daily cases than the rest of the states, but this has also declined over time (see negative sign of NY x date). Visualization is the most effective way to understand the trend according to the mobility indicators and date. Figure \ref{fig:prediction-plots} shows the prediction surfaces on three different dates: March 21, when the first states began implementing stay-at-home orders; April 5, fifteen days later; and April 20, fifteen more days later, and at a time when some states are considering relaxing stay-at-home orders. On March 21 there were still only minor departures from baseline mobility (recall that these are temporally lagged); the prediction surface at this point is relatively flat. This changes by April 5, when large increases in residential-based mobility are seen, along with greater variations in parks-based mobility. The prediction surface indicates an expectation of a greater number of new daily cases as both classes of mobility increase. On the last date examined, April 20, the trend becomes more steep for park-based mobility, even as this indicator continues to display large variations from the baseline in both directions. In general, higher levels of mobility tend to be associates with a higher number of new daily cases.

```{r model-estimation, include=FALSE}
#This is the regression model (forcing the regression through the intercept):
mod1 <- lm(log_new_cases ~  0 +
             day +
             day2 +
             parks2 + 
             parks + 
             parksXresidential +
             residential +
             residential2 +
             parksXday + 
             #parksXday2 +
             residentialXday +
             residentialXday2 +
             NY + 
             NYxday, 
           data = c19 #%>% filter(date >= "2020-04-01")
           )
summ(mod1)
```

```{r model-results, echo = FALSE}
data.frame(Variable = c("date",
                        "date^2",
                        "parks^2",
                        "parks",
                        "parks x residential",
                        "residential",
                        "residential^2",
                        "parks x date",
                        "residential x date",
                        "residential x date^2",
                        "NY",
                        "NY x date"),
           Coefficient = as.numeric(coefficients(mod1)),
           pval = as.numeric(summary(mod1)$coefficients[,4])) %>%
  mutate(pval = ifelse(pval >= 0.0001, round(pval, 4), "<0.0001")) %>%
  kable(#"latex",
        "html",
        booktabs = TRUE,
        digits = 4,
        caption = "\\label{tab:model-results}Results of estimating regression model. Dependent variable is log(New Cases + 0.0001).",
        align = c("l", "c", "c"),
        col.names = c("Variable", "Coefficient Estimate", "p-value")) %>%
  #kable_styling(latex_options = c("striped", "hold_position")) %>%
  footnote(general = c(paste("Coefficient of Determination $R^2$=", round(summary(mod1)$r.squared, 3)),
                       paste("Adjusted Coefficient of Determination $R^2$=", round(summary(mod1)$adj.r.squared, 3)),
                       paste("Standard Error $\\\\sigma$=", round(summary(mod1)$sigma, 3))),
           escape = FALSE)
```

>>>>

Antonio Paez (paezha@mcmaster.ca)  
School of Geography and Earth Science  
McMaster University  
1280 Main Street West, Hamilton, ON L8S 4K1  
Canada

(revised versin of paper after anonymous reviews submitted to Transport Findings on May 11, 2020)

@@

In 2020 Google released a set of Community Mobility Reports (GCMR). These reports are based on the company's location-tracking capabilities and measure changes in mobility with respect to a baseline. This novel source of data offers an opportunity to investigate potential correlations between mobility and transmission of COVID-19. Using data from the New York Times on COVID-19 cases and GCMR, this paper presents an analysis of mobility levels and new daily reported cases of COVID-19 by state in the US. The results provide insights about the utility and interpretability of GCMR for COVID-19 research and decision-making.

@@

The main policy tool to control the spread of the COVID-19 pandemic has been stay-at-home orders, which in the United States have been implemented on a state-by-state basis, with considerable variations in compliance. Concurrently, numerous initiatives have been developed to track the progress and the impact of the pandemic. As a result, there are new sources of data such as the recently-released Google Community Mobility Reports (GCMR)\footnote{https://www.google.com/covid19/mobility/}, as well as The New York Times repository of COVID-19 data\footnote{https://github.com/nytimes/covid-19-data}. These two open data sets offer novel opportunities to investigate in quasi-real time the relationship between mobility patterns and transmission of COVID-19.

@@

This paper is a reproducible research document. The source is an `R` markdown file available in a public repository\footnote{See folder Covid-19-Google-CMR-US in \url{https://github.com/paezha/Google-Mobility-Reports-and-COVID-19-US/}}.

@@

GCMR use aggregated and anonymized data to chart changes in mobility with respect to different classes of places (see Table \ref{tab:descriptive-statistics}). Mobility indicators are calculated based on the frequency and length of visits. The reports give percentage change from a baseline level, which corresponds to the median value of mobility of identical days of the week during the period between January 3 and Feb 6, 2020. Covid-19 data is compiled by The New York Times based on reports from state and local health agencies.

```{r descriptive-statistics, echo=FALSE}
data.frame(Variable = c("New Cases",
                        "date",
                        "retail",
                        "groceries",
                        "parks",
                        "transit",
                        "work",
                        "residential"),
           Definition = c("New Daily Cases of COVID-19",
                        "Date",
                        "Mobility trends for places like restaurants, cafes, shopping centers, theme parks, museums, libraries, and movie theaters",
                        "Mobility trends for places like grocery markets, food warehouses, farmers markets, specialty food shops, drug stores, and pharmacies",
                        "Mobility trends for places like local parks, national parks, public beaches, marinas, dog parks, plazas, and public gardens",
                        "Mobility trends for places like public transport hubs such as subway, bus, and train stations",
                        "Mobility trends for places of work",
                        "Mobility trends for places of residence"),
           min = c(as.character(min(c19$new_cases)),
                   as.character(min(c19$date)),
                   as.character(round(min(c19$retail), 2)),
                   as.character(round(min(c19$groceries), 2)),
                   as.character(round(min(c19$parks), 2)),
                   as.character(round(min(c19$transit), 2)),
                   as.character(round(min(c19$work), 2)),
                   as.character(round(min(c19$residential), 2))),
           median = c(as.character(median(c19$new_cases)),
                   as.character(median(c19$date)),
                   as.character(round(median(c19$retail), 2)),
                   as.character(round(median(c19$groceries), 2)),
                   as.character(round(median(c19$parks), 2)),
                   as.character(round(median(c19$transit), 2)),
                   as.character(round(median(c19$work), 2)),
                   as.character(round(median(c19$residential), 2))),
           max = c(as.character(max(c19$new_cases)),
                   as.character(max(c19$date)),
                   as.character(round(max(c19$retail), 2)),
                   as.character(round(max(c19$groceries), 2)),
                   as.character(round(max(c19$parks), 2)),
                   as.character(round(max(c19$transit), 2)),
                   as.character(round(max(c19$work), 2)),
                   as.character(round(max(c19$residential), 2))),
           sd = c(as.character(round(sd(c19$new_cases), 2)),
                   as.character(""),
                   as.character(round(sd(c19$retail), 2)),
                   as.character(round(sd(c19$groceries), 2)),
                   as.character(round(sd(c19$parks), 2)),
                   as.character(round(sd(c19$transit), 2)),
                   as.character(round(sd(c19$work), 2)),
                   as.character(round(sd(c19$residential), 2)))) %>%
  kable(#"latex",
        "html",
        booktabs = TRUE,
        caption = "\\label{tab:descriptive-statistics}Descriptive statistics of the data set") %>%
  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position")) %>%
  column_spec(2, width = "12em") %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

For analysis, all mobility indicators are centered so that the value of 1 is the base mobility, and a 0.01 deviation corresponds to a 1% change. The incubation time of the disease is between 2 and 12 days [95% interval; see @Lauer2020incubation]. Given this, it is to be expected that any changes in mobility will have a lagged effect (if any) on the discovery of new cases. For this reason, lagged moving averages of the mobility indicators are calculated. Furthermore, it is possible that mobility and reports of new cases of COVID-19 are endogenous, if the public adjust their mobility according to reports of the incidence. Therefore, in addition to being consistent with an incubation period, use of lagged indicators also helps to break this potential endogeneity.

The lagged indicators are calculated as the mean of the mobility indicator using the values from date-minus-12-days to date-minus-2-days. Furthermore, using the cumulative number of reported COVID-19 cases, the total number of new daily cases is calculated. This variable (log-transformed after adding a small constant) is paired with the corresponding lagged moving average of the mobility indicators. The log-transformation is useful to avoid negative values of daily new cases when making predictions. Table \ref{tab:descriptive-statistics} shows the descriptive statistics of the data set. Analysis is based on correlation analysis, multivariate regression, and data visualization.

@@

Table \ref{tab:correlation-analysis} shows that the mobility indicators are highly correlated with each other. Two variables are selected for multivariate analysis: parks- and work-related mobility. Work has a high correlation with the outcome variable, and its correlation with parks is relatively weak, which increases the information content of the two variables. Furthermore, parks- and work-related mobility represent two dimensions of out-of-home activities: mandatory and discretionary travel.

```{r check-correlations, echo=FALSE}
#Check the correlation between the mobility indicators and incidence:
c19 %>% 
  select(log_new_cases, retail, groceries, parks, transit, work, residential) %>%
  cor(use = "pairwise.complete.obs") %>%
  kable(#"latex",
        "html",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:correlation-analysis}Simple correlation between log(New Cases) and the mobility indicators")  %>%
  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position")) %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

A regression model is estimated with the log of new daily cases as the dependent variable. The covariates enter the regression in the form of a second order polynomial expansion. In addition, the date (centered on April 5) is introduced to account for the temporal trend of the pandemic. Finally, an indicator variable for the state of New York and an interaction with the date are used to distinguish the unusually high incidence of the disease there. The results appear in Table \ref{tab:model-results}. The model provides a good fit to the data and all variables reported are significant at $p<0.05$ or better.

There is an overall temporal trend that indicates a growing number of cases over time, at an accelerating rate (see positive sign of date^2). Mobility related to parks and to work both tend to increase the number of new cases; the negative sign for their interaction is indicative of the trade-offs between these two forms of mobility and their impact on new cases. The influence of parks-related mobility was relatively weaker early in the pandemic (negative sign of the parks x date term) but has become more important over time (positive sign of the parks x date^2 term). The opposite happens with work-related mobility, which started with a greater effect (positive sign of work x date term), but whose impact has declined over time (negative sign of work x date^2 term) New York has on average more new daily cases than the rest of the states, but this has declined over time (see negative sign of NY x date).

```{r model-estimation, include=FALSE}
#This is the regression model (forcing the regression through the intercept):
mod1 <- lm(log_new_cases ~  0 +
             day +
             day2 +
             parks2 + 
             parks + 
             parksXwork +
             work +
             work2 +
             parksXday + 
             parksXday2 +
             workXday +
             workXday2 +
             NY + 
             NYxday, 
           data = c19 #%>% filter(date >= "2020-04-01")
           )
summ(mod1)
```

```{r model-results, echo = FALSE}
data.frame(Variable = c("date",
                        "date^2",
                        "parks^2",
                        "parks",
                        "parks x work",
                        "work",
                        "work^2",
                        "parks x date",
                        "parks x date^2",
                        "work x date",
                        "work x date^2",
                        "NY",
                        "NY x date"),
           Coefficient = as.numeric(coefficients(mod1)),
           pval = as.numeric(summary(mod1)$coefficients[,4])) %>%
  mutate(pval = ifelse(pval >= 0.001, round(pval, 4), "<0.001")) %>%
  kable(#"latex",
        "html",
        booktabs = TRUE,
        digits = 4,
        caption = "\\label{tab:model-results}Results of estimating regression model. Dependent variable is log(New Cases + 0.0001).",
        align = c("l", "c", "c"),
        col.names = c("Variable", "Coefficient Estimate", "p-value")) %>%
  kable_styling(latex_options = c("striped", "HOLD_position")) %>%
  footnote(general = c(paste("Coefficient of Determination $R^2$=", round(summary(mod1)$r.squared, 3)),
                       paste("Adjusted Coefficient of Determination $R^2$=", round(summary(mod1)$adj.r.squared, 3)),
                       paste("Standard Error $\\\\sigma$=", round(summary(mod1)$sigma, 3))),
           escape = FALSE)
```

Visualization is the most effective way to understand the trend according to the mobility indicators and date. Figure \ref{fig:prediction-plots} shows the prediction surfaces on three different dates: March 21, when the first states began implementing stay-at-home orders; April 5, fifteen days later; and April 20, fifteen more days later, and at a time when some states started to consider relaxing stay-at-home orders.

On March 21 there were still only minor departures from baseline mobility (recall that these are temporally lagged); the prediction surface at this point is relatively flat. This changes by April 5, when work-based mobility has declined substantially. Although every state registers lower work-based mobility, there are large variations in parks-based mobility, with some states seeing increases of up to 60% for this class of mobility.

The prediction surface indicates an expectation of a greater number of new daily cases as either class of mobility increases, but the effect is not linear. On the last date examined, April 20, the trend becomes more steep for park-based mobility, even as this indicator continues to display large variations from the baseline in both directions. The white dashed lines in the plots are the folds of the saddles, and represent, for each date, the combination of parks- and work-related mobility levels that tended to minimize the emergence of new cases.
